Learning \strips\ action models with classical planning

### 2017-11-13 doc/icaps-18/strips-learning.tex
full version paper icaps18 with reorganized results

\maketitle
\begin{abstract}
This paper presents a novel approach for learning \strips\ action models from examples that compiles this inductive learning task into classical planning. Interestingly, the compilation approach is flexible to different amounts of available input knowledge; the learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states (no intermediate action or state is given). What is more, the compilation accepts partially specified action models and can be used to validate whether certain observations of plan executions follow a given \strips\ action model, even if this model is not fully specified.
\end{abstract}

\section{Introduction}
Besides {\em plan synthesis}~\cite{ghallab2004automated,geffner:book:2013}, planning action models are also useful for {\em plan/goal recognition}~\cite{ramirez2012plan}. At both planning tasks, an automated planner is required to reason about action models that correctly and completely capture the possible world transitions. Unfortunately, building planning action models is complex, even for planning experts, and this knowledge acquisition task is a bottleneck that limits the potential of planning~\cite{kambhampati:modellite:AAAI2007}.

On the other hand, Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models~\cite{fikes1971strips}, the vanilla action model for planning, is not straightforward though:
\begin{itemize}
\item The inputs to ML algorithms (the learning/training data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each plan possibly has different length).
\item The traditional output of ML algorithms is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of preconditions, negative and positive effects, that define the possible state transitions.
\end{itemize}

Learning \strips\ action models is a well-studied problem with sophisticated algorithms, like {\sc ARMS}~\cite{yang2007learning}, {\sc SLAF}~\cite{amir:alearning:JAIR08} or {\sc LOCM}~\cite{cresswell2013acquiring} that do not require full knowledge of the intermediate states traversed by the example plans. Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. The compilation approach is appealing itself because it opens the door to the bootstrapping of planning action models but also because:
\begin{enumerate}
\item Is flexible to different amounts of available input knowledge. The learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states where no intermediate state  or action is observed.
\item Accepts previous knowledge about the structure of the actions in the form of partially specified action models. In the extreme, the compilation could be used to validate whether observed plan executions are valid for a given \strips\ action model.
\end{enumerate}

The first section of the paper presents the classical planning model, its extension to conditional effects (which is a requirement of the proposed compilation) and formalizes the \strips\ action model (the output of the addressed learning task). The second section formalizes the learning of \strips\ action models with regard to different amounts of available input knowledge. The third and fourth sections describe our approach for addressing the formalized learning tasks. Finally, the last sections report the data collected during the empirical evaluation of our approach, discuss the strengths and weaknesses of our approach and propose several opportunities for future research.

@@

A {\em state} $s$ is a total assignment of values to fluents, i.e. $|s|=|F|$, so the size of the state space is $2^{|F|}$. Explicitly including negative literals $\neg f$ in states simplifies subsequent definitions but often, we will abuse notation by defining a state $s$ only in terms of the fluents that are true in $s$, as is common in \strips\ planning.

A {\em classical planning frame} is a tuple $\Phi=\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions. Each action $a\in A$ comprises three sets of literals:
\begin{itemize}
\item $\pre(a)\subseteq\mathcal{L}(F)$, called {\em preconditions}, that defines the literals that must hold for the action to be applicable.
\item $\eff^+(a)\subseteq\mathcal{L}(F)$, called {\em positive effects}, that defines the fluents set to true by the action application.
\item $\eff^-(a)\subseteq\mathcal{L}(F)$, called {\em negative effects}, that defines the fluents set to false by the application of the action.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in state $s$ iff $\pre(a)\subseteq s$, and the result of applying $a$ in $s$ is the {\em successor state} $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\subseteq\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces a state sequence $\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.~if the goal condition is satisfied at the last state reached after following the application of $\pi$ in $I$. We denote with $|\pi|$ the {\em plan length}.

\subsection{Classical planning with conditional effects}
Our approach for leaning \strips\ action models is compiling this leaning task into a classical planning task with conditional effects. Conditional effects allow us to compactly define actions whose effects depend on the current state. Many classical planners cope with conditional effects without compiling them away. In fact, supporting conditional effects is a requirement of IPC-2014~\cite{vallati:IPC:AIM2015} and IPC-2018.

Now an action $a\in A$ has a set of literals $\pre(a)\in\mathcal{L}(F)$, called the {\em precondition}, and a set of {\em conditional effects} $\cond(a)$. Each conditional effect $C\rhd E\in\cond(a)$ is composed of two sets of literals $C\in\mathcal{L}(F)$, the {\em condition}, and $E\in\mathcal{L}(F)$, the {\em effect}.

An action $a\in A$ is {\em applicable} in state $s$ if and only if $\pre(a)\subseteq s$, and the resulting set of {\em triggered effects} is
\[
triggered(s,a)=\bigcup_{C\rhd E\in\cond(a),C\subseteq s} E,
\]
i.e.~effects whose conditions hold in $s$.

The result of applying action $a$ in a state $s$ is the {\em successor} state $\theta(s,a)=\{s\setminus\eff_c^-(s,a))\cup\eff_c^+(s,a)\}$ where $\eff_c^-(s,a)\subseteq triggered(s,a)$ and $\eff_c^+(s,a)\subseteq triggered(s,a)$ are the triggered negative and positive effects, respectively.

\subsection{The \strips\ action schema and the {\em variable} objects}
This work addresses the learning of PDDL action schemes that follow the \strips\ requirement~\cite{mcdermott1998pddl,fox2003pddl2}. Figure~\ref{fig:stack} shows the schema that corresponds to the {\em stack} action from a four-operator {\em blocksworld}~\cite{slaney2001blocks}.

\begin{figure}
\begin{footnotesize}
\begin{verbatim}
(:action stack
  :parameters (?x1 ?x2 - object)
  :precondition (and (holding ?x1) (clear ?x2))
  :effect (and (not (holding ?x1))
               (not (clear ?x2))
               (handempty) (clear ?x1)
               (on ?x1 ?x2)))
\end{verbatim}
\end{footnotesize}
 \caption{\small Example of a \strips\ operator schema that corresponds to the {\em stack} action from the {\em blocksworld} represented in PDDL.}
\label{fig:stack}
\end{figure}

To formalize the output of the learning task, we assume that there exists a set of {\em predicates} $\Psi$ and that fluents $F$ are instantiated from these predicates, as in PDDL. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of objects $\Omega$, the set of fluents $F$ is then induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^k$ is the $k$-th Cartesian power of $\Omega$. Likewise, we assume that actions $a\in A$ are instantiated from \strips\ operator schemes.

Let $\Omega_v=\{v_i\}_{i=1}^{\operatorname*{max}_{a\in A} ar(a)}$ be a new set of objects denoted as {\em variable names}, $\Omega\cap\Omega_v=\emptyset$, and that is bound by the maximum arity of an action in a given planning frame. For instance, in a three-block blocksworld $\Omega=\{block_1, block_2, block_3\}$ while $\Omega_v=\{v_1, v_2\}$ because the operators with the maximum arity, {\small\tt stack} and {\small\tt unstack}, have two parameters each.

Let us define also a new set of fluents $F_{v}$, $F\cap F_v=\emptyset$, that results from instantiating $\Psi$ using only {\em variable objects} in $\Omega_v$. This set defines the elements that can appear in an action schema. In the blocksworld $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$),on($v_1,v_2$), on($v_2,v_1$),on($v_2,v_2$)\}}.

We are now ready to define $\Xi$, a set of operator schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ such that:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, is the operator {\em header} defined by its name and $pars(\xi)=\{v_i\}_{i=1}^{ar(\xi)}$, an enumeration of the variable objects bound by the operator arity. The headers for a 4-operator blocksworld are: {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the negative effects $del(\xi)\subseteq F_v$, and the positive effects $add(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}

\section{Learning \strips\ action models}
Learning \strips\ action models from fully available input knowledge, i.e. from plans where the {\em pre-} and {\em post-states} of every action in a plan are available, is straightforward. In this case, because intermediate states are available, \strips\ operator schemes are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals that appears in all the pre-states that correspond to the same operator schema.

@@

\subsubsection{Learning from (initial, final) state pairs.} This learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions. No information about the actions in the plans is given. This learning task is formalized as $\Lambda=\tup{\Psi,\Sigma}$:
\begin{itemize}
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning domain.
\item $\Sigma=\{\sigma_1,\ldots,\sigma_{\tau}\}$ is a set of labels s.t. each label $\sigma_t=(s_0^t,s_{n}^t)$, {\tt\small $1\leq t\leq \tau$}, is an $(initial, final)$ state pair that comprises the {\em final} state $s_{n}^t$ resulting from executing an unknown plan $\pi_t$ starting from a given {\em initial} state $s_0^t$.
\end{itemize}

\subsubsection{Learning from labeled plans.}
Here we augment the amount of provided input knowledge with the actions executed by the observed agent and define the learning task as $\Lambda'=\tup{\Psi,\Sigma,\Pi}$:

\begin{itemize}
\item $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is a given set of example plans where each plan $\pi_t=\tup{a_1^t, \ldots, a_n^t}$, {\small $1\leq t\leq \tau$}, is an action sequence that induces the corresponding state sequence $\tup{s_0^t, s_1^t, \ldots, s_n^t}$ such that, for each {\small $1\leq i\leq n$}, $a_i^t$ is applicable in the state $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
\end{itemize}

@@

@@

\node (goalA) [draw, right=10mm of initD]{A};
\node (goalB) [draw, right=10mm of initC]{B};
\node (goalC) [draw, right=10mm of initB]{C};
\node (goalD) [draw, right=10mm of initA]{D};
\end{tikzpicture}
\vspace{0.6cm}
\end{subfigure}%
 \caption{\small Example of a task for learning a \strips\ action model in the blocksworld.}
\label{fig:lexample}
\end{figure}

\subsubsection{Learning from partially specified action models.}
We may not require to start learning the \strips\ action models from scratch so we augment here the inputs of the learning task with partially specified operator schemes. In this case the leaning task is defined as $\Lambda''=\tup{\Psi,\Sigma,\Pi,\Xi_0}$ where:
\begin{itemize}
\item $\Xi_0$ is a partially specified action model in which some preconditions and effects are a priori known.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi$ that is compliant just with the predicates in $\Psi$, and the given set of initial and final states $\Sigma$. In this learning scenario, a solution must not only determine a possible \strips\ action model but also the plans $\pi_t$, {\tt\small $1\leq t\leq \tau$} that explain the given labels $\Sigma$ using the learned \strips\ model. A solution to $\Lambda'$ is a set of \strips\ operator schema $\Xi$ (with one schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ for each action with a different name in the example plans $\Pi$) compliant with the predicates in $\Psi$, the example plans $\Pi$, and their corresponding labels $\Sigma$.  Finally a solution to $\Lambda''$ is a set of \strips\ operator schema $\Xi$ compliant as well with the provided partially specified action model $\Xi_0$.

\section{Learning \strips\ action models with planning}
Our approach for addressing a learning task $\Lambda$, $\Lambda'$ or $\Lambda''$, is compiling it into a classical planning task with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the \strips\ action model $\Xi$. The solution plan has a {\em prefix} that, for each $\xi\in\Xi$, determines the fluents from $F_v$ that belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the programmed \strips\ action model $\Xi$ using the given input knowledge (the set of labels $\Sigma$, and $\Pi$ and/or $\Xi_0$ if available). For every label $\sigma_t\in \Sigma$, the solution plan has a postfix that produces a final state $s_{n}^t$ starting from the corresponding initial state $s_0^t$ using the programmed action model $\Xi$. We call this process the validation of the programmed \strips\ action model $\Xi$, at the learning example {\small $1\leq t\leq \tau$}. %If information about the corresponding plan $\pi_t\in \Pi$ is available, then it is also used in the validation.
\end{enumerate}

To formalize our compilations we first define {\small $1\leq t\leq \tau$} classical planning instances $P_t=\tup{F,\emptyset,I_t,G_t}$ that belong to the same planning frame (i.e. same fluents and actions and differ only in the initial state and goals). The fluents $F$ are built instantiating the predicates in $\Psi$ with the objects appearing in the input labels $\Sigma$. Formally $\Omega=\{o|o\in \bigcup_{\small 1\leq t\leq \tau} obj(s_0^t)\}$, where $obj$ is a function that returns the set of objects that appear in a fully specified state. The set of actions, $A=\emptyset$, is empty because the action model is initially unknown. Finally, the initial state $I_t$ is given by the state $s_0^t\in \sigma_t$ while goals $G_t$, are defined by the state $s_n^t\in \sigma_t$.

Now we are ready to formalize the compilations. We start with $\Lambda$, because this learning task requires the less input knowledge. Given a learning task $\Lambda=\tup{\Psi,\Sigma}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with:
\begin{itemize}
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v$ and $\xi \in \Xi$. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi$. For instance in the working example the preconditions of the $stack$ schema are represented by the fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to true.
\item A fluent $mode_{prog}$ indicating whether the operator schemes are being programmed or they are being validated (already programmed).
\item Fluents $\{test_t\}_{1\leq t\leq \tau}$, indicating the learning example where the programmed action model is being validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0^1$ (the initial state of the first learning example), every $pre_f(\xi)\in F_{\Lambda}$ (our compilation assumes that initially any operator schema is programmed with every possible precondition, no negative effect and no positive effect) and $mode_{prog}$ set to true.
\item $G_{\Lambda}=\bigcup_{1\leq t\leq \tau}\{test_t\}$, indicates that the programmed action model is validated in all the learning examples.
\item $A_{\Lambda}$ contains actions of three different kinds:
\begin{enumerate}
\item The actions for {\em programming} an operator schema $\xi\in\Xi$, which includes:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\bf adding} a {\em negative} or a {\em positive} effect $f\in F_v$ to the action schema $\xi\in\Xi$.

>>>>

\maketitle
\begin{abstract}
This paper presents a novel approach for learning \strips\ action models from examples that compiles this inductive learning task into classical planning. Interestingly, the compilation approach is flexible to different amounts of available input knowledge; the learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states (no intermediate action or state is given). What is more, the compilation accepts partially specified action models and can be used to validate whether the observation of a plan execution follows a given \strips\ action model, even if this model is not fully specified.
\end{abstract}

\section{Introduction}
Besides {\em plan synthesis}~\cite{ghallab2004automated}, planning action models are also useful for {\em plan/goal recognition}~\cite{ramirez2012plan}. At both planning tasks, an automated planner is required to reason about action models that correctly and completely capture the possible world transitions~\cite{geffner:book:2013}. Unfortunately, building planning action models is complex, even for planning experts, and this knowledge acquisition task is a bottleneck that limits the potential of AI planning~\cite{kambhampati:modellite:AAAI2007}.

On the other hand, Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models, the vanilla action model for planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The {\em input} to ML algorithms (the learning/training data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each plan possibly has a different length).
\item The {\em output} of ML algorithms usually is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of preconditions, negative and positive effects, that define the possible state transitions.
\end{itemize}

Learning \strips\ action models is a well-studied problem with sophisticated algorithms, like {\sc ARMS}~\cite{yang2007learning}, {\sc SLAF}~\cite{amir:alearning:JAIR08} or {\sc LOCM}~\cite{cresswell2013acquiring} that do not require full knowledge of the intermediate states traversed by the example plans. Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. The compilation approach is appealing by itself because opens the door to the bootstrapping of planning action models but also because:
\begin{enumerate}
\item Is flexible to different amounts of available input knowledge. The learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states where no intermediate state  or action is observed.
\item Accepts previous knowledge about the structure of the actions in the form of partially specified action models. In the extreme, the compilation can validate whether an observed plan execution is valid for a given \strips\ action model, even if this model is not fully specified.
\end{enumerate}

The second section of the paper formalizes the classical planning model, its extension to {\em conditional effects} (a requirement of the proposed compilation) and the \strips\ action model (the output of the addressed learning task). The third section formalizes the learning of \strips\ action models with regard to different amounts of available input knowledge. The fourth and fifth sections describe our approach for addressing the formalized learning tasks. Finally, the last sections report the data collected in a empirical evaluation, discuss the strengths and weaknesses of our approach and propose several opportunities for future research.

@@

A {\em state} $s$ is a full assignment of values to fluents, i.e. $|s|=|F|$, so the size of the state space is $2^{|F|}$. Explicitly including negative literals $\neg f$ in states simplifies subsequent definitions but often, we will abuse notation by defining a state $s$ only in terms of the fluents that are true in $s$, as is common in \strips\ planning.

A {\em classical planning frame} is a tuple $\Phi=\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions. Each action $a\in A$ comprises three sets of literals:
\begin{itemize}
\item $\pre(a)\subseteq\mathcal{L}(F)$, called {\em preconditions}, the literals that must hold for the action $a\in A$ to be applicable.
\item $\eff^+(a)\subseteq\mathcal{L}(F)$, called {\em positive effects}, that defines the fluents set to true by the application of the action $a\in A$.
\item $\eff^-(a)\subseteq\mathcal{L}(F)$, called {\em negative effects}, that defines the fluents set to false by the action application.
\end{itemize}
We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\subseteq\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces a state sequence $\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. We denote with $|\pi|$ the {\em plan length}. A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.~if the goal condition is satisfied at the last state reached after following the application of $\pi$ in $I$.

\subsection{Classical planning with conditional effects}
Our approach for learning \strips\ action models is compiling this leaning task into a classical planning task with conditional effects. Conditional effects allow us to compactly define actions whose effects depend on the current state. Supporting conditional effects is now a requirement of the IPC~\cite{vallati:IPC:AIM2015} and many classical planners cope with conditional effects without compiling them away.

An action $a\in A$ has now a set of {\em preconditions} $\pre(a)\in\mathcal{L}(F)$ and a set of {\em conditional effects} $\cond(a)$. Each conditional effect $C\rhd E\in\cond(a)$ is composed of two sets of literals $C\in\mathcal{L}(F)$, the {\em condition}, and $E\in\mathcal{L}(F)$, the {\em effect}.

An action $a\in A$ is {\em applicable} in a state $s$ if and only if $\pre(a)\subseteq s$, and the resulting set of {\em triggered effects} are the effects whose conditions hold in $s$:
\[
triggered(s,a)=\bigcup_{C\rhd E\in\cond(a),C\subseteq s} E,
\]

The result of applying an action $a$ in a state $s$ is the {\em successor} state $\theta(s,a)=\{s\setminus\eff_c^-(s,a))\cup\eff_c^+(s,a)\}$ where $\eff_c^-(s,a)\subseteq triggered(s,a)$ and $\eff_c^+(s,a)\subseteq triggered(s,a)$ are the triggered {\em negative} and {\em positive} effects, respectively.

\subsection{\strips\ action schemes and {\em variable name} objects}
This work addresses the learning of PDDL action schemes that follow the \strips\ requirement~\cite{mcdermott1998pddl,fox2003pddl2}. Figure~\ref{fig:stack} shows the schema, coded in PDDL, for the {\em stack} action from a four-operator {\em blocksworld}~\cite{slaney2001blocks}.

\begin{figure}
\begin{footnotesize}
\begin{verbatim}
(:action stack
  :parameters (?v1 ?v2 - object)
  :precondition (and (holding ?v1) (clear ?v2))
  :effect (and (not (holding ?v1))
               (not (clear ?v2))
               (handempty) (clear ?v1)
               (on ?v1 ?v2)))
\end{verbatim}
\end{footnotesize}
 \caption{\small \strips\ operator schema coding, in PDDL, the {\em stack} action from the {\em blocksworld}.}
\label{fig:stack}
\end{figure}

To formalize the output of the learning task, we assume that fluents $F$ are instantiated from a set of {\em predicates} $\Psi$, as in PDDL. Each predicate $p\in\Psi$ has an argument list of arity $ar(p)$. Given a set of {\em objects} $\Omega$, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

Let $\Omega_v=\{v_i\}_{i=1}^{\operatorname*{max}_{a\in A} ar(a)}$ be a new set of objects $\Omega\cap\Omega_v=\emptyset$, denoted as {\em variable names}, and that is bound by the maximum arity of an action in a given planning frame. For instance, in a three-block blocksworld $\Omega=\{block_1, block_2, block_3\}$ while $\Omega_v=\{v_1, v_2\}$ because the operators with the maximum arity, {\small\tt stack} and {\small\tt unstack}, have two parameters each.

Let us also define $F_v$, a new set of fluents $F\cap F_v=\emptyset$, that results from instantiating $\Psi$ using only the objects in $\Omega_v$ and that defines the elements that can appear in an action schema. For instance, in the blocksworld, $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$), on($v_1,v_2$), on($v_2,v_1$), on($v_2,v_2$)\}}.

Finally, we assume that actions $a\in A$ are instantiated from \strips\ operator schemes $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ where:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, is the operator {\em header} defined by its name and corresponding {\em variable names}, $pars(\xi)=\{v_i\}_{i=1}^{ar(\xi)}$. For instance, the headers for a four-operator blocksworld are: {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the negative effects $del(\xi)\subseteq F_v$, and the positive effects $add(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}

\section{Learning \strips\ action models}
Learning \strips\ action models from fully available input knowledge, i.e. from plans where the {\em pre-} and {\em post-states} of every action in a plan are available, is straightforward. When any intermediate state is available, \strips\ operator schemes are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Preconditions are derived lifting the minimal set of literals that appears in all the pre-states of the corresponding actions.

@@

\subsubsection{Learning from (initial, final) state pairs.} This learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions. No information about the actions in the plans is given. This learning task is formalized as $\Lambda=\tup{\Psi,\Sigma}$:
\begin{itemize}
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning domain.
\item $\Sigma=\{\sigma_1,\ldots,\sigma_{\tau}\}$ is a set of $(initial, final)$ state pairs, that we call {\em labels}. Each label $\sigma_t=(s_0^t,s_{n}^t)$, {\tt\small $1\leq t\leq \tau$}, comprises the {\em final} state $s_{n}^t$ resulting from executing an unknown plan $\pi_t$ starting from the {\em initial} state $s_0^t$.
\end{itemize}

\subsubsection{Learning from labeled plans.}
Here we augment the input knowledge with the actions executed by the observed agent and define the learning task $\Lambda'=\tup{\Psi,\Sigma,\Pi}$:

\begin{itemize}
\item $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is a given set of example plans where each plan $\pi_t=\tup{a_1^t, \ldots, a_n^t}$, {\small $1\leq t\leq \tau$}, is an action sequence that induces the corresponding state sequence $\tup{s_0^t, s_1^t, \ldots, s_n^t}$ such that, for each {\small $1\leq i\leq n$}, $a_i^t$ is applicable in $s_{i-1}^t$ and generates $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
\end{itemize}

@@

\subsubsection{Learning from partially specified action models.}
We may not require to start learning from scratch so we augment the learning input with partially specified operator schemes. This leaning task is defined as $\Lambda''=\tup{\Psi,\Sigma,\Pi,\Xi_0}$:
\begin{itemize}
\item $\Xi_0$ is a partially specified action model in which some preconditions and effects are a priori known.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi$ that is compliant just with the predicates in $\Psi$, and the given set of initial and final states $\Sigma$. In this learning scenario, a solution must not only determine a possible \strips\ action model but also the plans $\pi_t$, {\tt\small $1\leq t\leq \tau$} that explain the given labels $\Sigma$ using the learned \strips\ model. A solution to $\Lambda'$ is a set of \strips\ operator schema $\Xi$ (one schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ for each action with a different name in the example plans $\Pi$) compliant with the predicates in $\Psi$, the example plans $\Pi$, and their corresponding labels $\Sigma$.  Finally a solution to $\Lambda''$ is a set of \strips\ operator schema $\Xi$ compliant as well with the provided partially specified action model $\Xi_0$.

@@

\node (goalA) [draw, right=10mm of initD]{A};
\node (goalB) [draw, right=10mm of initC]{B};
\node (goalC) [draw, right=10mm of initB]{C};
\node (goalD) [draw, right=10mm of initA]{D};
\end{tikzpicture}
\vspace{0.6cm}
\end{subfigure}%
 \caption{\small Example of a task for learning a \strips\ action model in the blocksworld from a single labeled plan.}
\label{fig:lexample}
\end{figure}

\section{Learning \strips\ action models with planning}
Our approach for addressing a learning task $\Lambda$, $\Lambda'$ or $\Lambda''$, is compiling it into a classical planning task with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the \strips\ action model $\Xi$. A solution plan has a {\em prefix} that, for each $\xi\in\Xi$, determines the fluents from $F_v$ that belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the programmed \strips\ action model $\Xi$ in the given input knowledge (the labels $\Sigma$, and $\Pi$ and/or $\Xi_0$ if available). For every label $\sigma_t\in \Sigma$, a solution plan has a postfix that produces a final state $s_{n}^t$ starting from the corresponding initial state $s_0^t$ using the programmed action model $\Xi$. We call this process the validation of the programmed \strips\ action model $\Xi$, at the learning example {\small $1\leq t\leq \tau$}. %If information about the corresponding plan $\pi_t\in \Pi$ is available, then it is also used in the validation.
\end{enumerate}

To formalize our compilation we first define {\small $1\leq t\leq \tau$} classical planning instances $P_t=\tup{F,\emptyset,I_t,G_t}$ that belong to the same planning frame (i.e. same fluents and actions but differ in the initial state and goals). Fluents $F$ are built instantiating the predicates in $\Psi$ with the objects appearing in the input labels $\Sigma$. Formally $\Omega=\{o|o\in \bigcup_{\small 1\leq t\leq \tau} obj(s_0^t)\}$, where $obj$ is a function that returns the set of objects that appear in a fully specified state. The set of actions, $A=\emptyset$, is empty because the action model is initially unknown. Finally, the initial state $I_t$ is given by the state $s_0^t\in \sigma_t$ while goals $G_t$, are defined by the state $s_n^t\in \sigma_t$.

Now we are ready to formalize the compilations. We start with $\Lambda$, because it requires less input knowledge. Given a learning task $\Lambda=\tup{\Psi,\Sigma}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ extends $F$ with:
\begin{itemize}
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v$ and $\xi \in \Xi$. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$}.
\item A fluent $mode_{prog}$ indicating whether the operator schemes are being programmed or validated (already programmed) and fluents $\{test_t\}_{1\leq t\leq \tau}$, indicating the example where the action model is being validated. 
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0^1$ (the initial state of the first label), every $pre_f(\xi)\in F_{\Lambda}$ and $mode_{prog}$ set to true. Our compilation assumes that initially any operator schema is programmed with every possible precondition, no negative effect and no positive effect.
\item $G_{\Lambda}=\bigcup_{1\leq t\leq \tau}\{test_t\}$, indicates that the programmed action model is validated in all the learning examples.
\item $A_{\Lambda}$ contains actions of three kinds:
\begin{enumerate}
\item Actions for {\em programming} an operator schema $\xi\in\Xi$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\bf adding} a {\em negative} or {\em positive} effect $f\in F_v$ to the action schema $\xi\in\Xi$.
